A tutorial on Generalised Additive Mixed Effects Models for bilingualism research

### 2023-12-09 manuscript/manuscript.qmd

```{r}
#| label: fig-d1-dag
#| layout-ncol: 2
#| echo: false

@@

```{r}
gam_2_preds <- predict_gam(
  gam_2, length_out = 25,
  series = "proficiency",
  exclude_terms = c("s(subj)", "s(subj,proficiency)")
)

@@

```{r}
gam_2_preds_t <- predict_gam(
  gam_2, length_out = 25,
  series = "time",
  exclude_terms = c("s(subj)", "s(subj,proficiency)")
)

@@

@@

```{r}
dat2
```

@@

To obviate this limitation, we can simply construct a new predictor with the combination of vowel and group and use that as the `by`-variable in the smooth terms.

```{r}
#| label: gam-3

dat2 <- dat2 %>%
  mutate(
    group_vowel = interaction(group, vowel)
  )

gam_3 <- gam(
  eud ~
    s(timep, by = group_vowel, k = 9) +
    s(timep, subj, by = group_vowel, bs = "fs", m = 1),
  data = dat2
)
```

>>>>

```{r}
#| label: fig-d1-dag
#| fig-cap: "Dyrected Acyclic Graphs for time, proficiency and learning score."
#| fig-subcap:
#| - "No direct effect of time on learning scores."
#| - "Direct effect of time on learning scores."
#| layout-ncol: 2
#| fig-width: 3
#| fig-height: 1.5
#| echo: false

@@

```{r}
#| label: gam-2-preds

gam_2_preds <- predict_gam(
  gam_2, length_out = 25,
  series = "proficiency",
  exclude_terms = c("s(subj)", "s(subj,proficiency)")
)

@@

```{r}
#| label: gam-2-preds-t

gam_2_preds_t <- predict_gam(
  gam_2, length_out = 25,
  series = "time",
  exclude_terms = c("s(subj)", "s(subj,proficiency)")
)

@@

[LIT REVIEW]

@@

```{r}
#| label: dat2

dat2
```

@@

To obviate this limitation, we can simply construct a new predictor with the combination of vowel and group and use that as the `by`-variable in the smooth terms.
Due to how smooth terms are constructed when you include a `by`-variable, it is necessary to also include the variable as a parametric term (i.e. a classical linear term).

```{r}
#| label: dat2-prep

dat2 <- dat2 %>%
  mutate(
    vowel = as.factor(vowel),
    group_vowel = interaction(group, vowel),
  )
```

```{r}
#| label: gam-3
#| eval: false

gam_3 <- gam(
  eud ~
    # Parametric term for group_vowel
    group_vowel +
    # Smooth term with group_vowel by-variable
    s(timep, by = group_vowel, k = 9) +
    # Random factor smooths
    s(timep, subj, bs = "fs", m = 1),
  data = dat2
)
```

```{r}
#| label: gam-3-eval
#| echo: false
gam_3_path <- "data/gam_3.rds"
if (file.exists(gam_3_path)) {
  gam_3 <- readRDS(gam_3_path)
} else {
  gam_3 <- gam(
    eud ~
      # Parametric term for group_vowel
      group_vowel +
      # Smooth term with group_vowel by-variable
      s(timep, by = group_vowel, k = 9) +
      # Random factor smooths
      s(timep, subj, bs = "fs", m = 1),
    data = dat2
  )
  saveRDS(gam_3, gam_3_path)
}
```

```{r}
#| label: gam-3-sum

summary(gam_3)
```

Let's extract the model predictions and inspect them.

```{r}
#| label: gam-3-preds

gam_3_preds <- predict_gam(gam_3, exclude_terms = "s(timep,subj)")
gam_3_preds
```

The function `predict_gam()` automatically samples 10 equidistant points from numeric variables (like `timep`, based on the default `length_out = 10` argument) and all levels in categorical variables (like `group_vowel`).
For each combination of the sampled points and levels, the function returns the value of the outcome (`eud`) and the standard error (`se`). The lower (`lower_ci`) and upper (`upper_ci`) boundaries of the 95% Confidence Interval are also returned.

Furthermore, `predict_gam()` returns an object of class `tidygam` which can be plotted with `plot()` (similarly to how `ggpredict()` from the ggeffects package works, XXX).
However, we might want to do some processing of the predictions before plotting: in particular, we might want to split the `group_vowel` column into the two original variables, `group` and `vowel`. We can achieve this straight from the function. And we might also want to extract more than 10 time points, to get a smoother curve.

```{r}
#| label: gam-3-preds-2

gam_3_preds_2 <- predict_gam(
  gam_3,
  length_out = 25,
  exclude_terms = "s(timep,subj)",
  separate = list(group_vowel = c("group", "vowel"))
)
gam_3_preds_2
```

The syntax `list(group_vowel = c("group", "vowel"))` instructs the function to split the `group_vowel` column into two columns, `group` and `vowel`.
Splitting is done based on the `sep_by` argument, which is `.` by default.
Now we can plot the predictions.

```{r}
#| label: fig-gam-3
#| fig-cap: "Predicted Euclidean distance along the duration of the vowel, for /a, i, u/ in simultaneous bilinguals (es), beginners late bilinguals (es_0) and advanced late bilinguals (es_1)"

gam_3_preds_2 %>%
  plot(series = "timep", comparison = "group") +
  facet_grid(cols = vars(vowel))
```
